Interpolation and extrapolation based performance measurement of Analytical and Artificial Neural Network models in predicting hot deformation behavior of a medium carbon steel

### 2022-12-28 Article.tex
PriorEditionVersion

\title{\textcolor{red}{Interpolation and extrapolation based performance measurement of Analytical and Artificial Neural Network models in predicting hot deformation behavior of a medium carbon steel}}

@@

\begin{abstract}
\textcolor{purple}{An analysis of the use of analytical models and Artificial Neural Networks is proposed in this paper to study the high-temperature deformation behavior of medium carbon steel alloy during hot compression tests, on a Gleeble thermomechanical simulator, over a wide range of strains ($0.0-0.7$), strain rates ($0.001~\ps-5~\ps$) and temperatures ($1050\celsius-1250\celsius$). From the experimental data (stress-true strain curves), 5 analytical models such as Johnson-Cook, Modified Zerilli-Armstrong, Hansel-Spittel, Arrhenius and PTM on the one hand, and an Artificial Neural Networks model with 3 inputs, 2 hidden layers of 15 and 7 neurons respectively and 1 output, on the other hand, have been identified to predict the flow stress $\sigma^y$ of this alloy. The comparison between the analytical models, the Artificial Neural Networks model and the experimental results, carried out on the basis of the Mean Absolute Relative Error ($\AARE$) and the Root Mean Square Error ($\RMSE$), were used to quantify these models. It is shown that the Johnson-Cook and  Modified Zerilli-Armstrong models are not suitable to predict the high temperature behavior of this medium carbon steel alloy, while the Hansel-Spittel, PTM and Arrhenius models can be used to predict the behavior of this alloy with an acceptable error. On the other hand, the Artificial Neural Networks model results show excellent agreement between the predicted flow stresses $\sigma^y$ and experiment with comparatively significantly lower error. To validate the performance of these models, their ability to interpolate and extrapolate the experimental data was also tested. The HS, PTM and AR analytical models have shown a good interpolation and extrapolation capacity with a notable superiority for the AR model compared to the other two. The Artificial Neural Networks model is still superior to the AR model from the point of view of its global behavior and its performance}. \textcolor{red}{An approach for a better usage of these models on unknown data is proposed where each constitutive law must be multiplied by a factor $\lambda$ only for models whose correlation with experience is acceptable}.
\end{abstract}

@@

%----------------------------------------------------------------------------------
\section{Introduction\label{sec:Introduction}}
%----------------------------------------------------------------------------------
\textcolor{purple}{Due to their excellent weldability, high machinability, good wear resistance, medium carbon steels nowadays are widely used in industrial fields. In addition, the hardening of these alloys through a heat treatment process and their high forgeability are very important characteristics that can be used for manufacturing large structures. As application, many automotive parts for steering, engines, suspensions and transmissions are made from these steels. Indeed, the medium carbon steels come to substitute quench-temper steels in different automative parts previously used as connecting rods produced by hot forging of low carbone steel alloys \cite{murugesan2019two, murugesan2019hybrid, tavakoli2019ferrite, ebrahimi2017flow}. Several metallurgical phenomena such as work hardening (WH) and dynamic softening (dynamic recovery (DRV) and recrystallization (DRX)), which very often occur during hot deformation, influence the microstructure evolution of materials during hot deformation. The processes involved in hot deformation are highly critical to the quality and workability of the final product \cite{Ashtiani-2012}. In order to estimate the influence of the different parameters involved in hot deformation before the experiment, numerical simulation like Finite Element Method (FEM) is widely used \cite{He-2013, Changizian-2012}. For example, FEM has been used to simulate and analyze hot forming processes of steel, such as forging, rolling, extrusion and casting, but the quality of the numerical solution provided by this method is highly dependent on the constitutive equation of the modeled material \cite{Mandal-2009, Qin-2010, Ji-2018}. To this end, a mathematical formulation is typically used to represent the flow stress $\sigma^y$ of a material from the strain $\varepsilon$, the strain rate $\mdot\varepsilon$ and the temperature $T$ in a form that can be used in FEM software \cite{Lin-2008-P, Wu-2012}}.

\textcolor{purple}{Various forms of constitutive equations have been proposed to describe the flow stress of materials, among which, phenomenological, semi-empirical, physical models or models based on Artificial Neural Networks (ANN) \cite{Rusinek-2010, Shin-2010, Lin-2011, Pantale-2021}. Among the large number of models available in the literature, the Johnson-Cook (JC) \cite{Johnson-1983} and Zerilli-Armstrong (ZA) \cite{Zerilli-1987} models are the best known and most widely available in FEM codes.
The Johnson-Cook model is the most widely used because it is simple to identify and use and has few parameters to determine \cite{NematNasser-2003, Khan-2004}.
However, this model, in its original state, has some shortcomings related to nonlinear coupling phenomena not taken into account in the material behavior.
We refer here to strain hardening, strain rate hardening and thermal effects that are taken into account in a non-coupled manner in the JC model as analyzed by Jia \eal \cite{Jia-2021}. To circumvent these shortcomings, several modified forms of the Johnson-Cook model have been developed in the past \cite{Rule-1998, Vural-2003, Lin-2010, Lin-2012, Li-2013, Zhang-2015, Zhou-2019}.
Due to its more physics-based formulation, the ZA model is often preferred to the JC model because it allows the coupling of strain rate and temperature effects \cite{Johnson-1988, Voyiadjis-2005, Dey-2007}.
However, this model has other drawbacks related to the limit value of the test temperature. Indeed, this model applies only to low strain rates and for temperatures limited to about 60~\% of $T_m$ (where $T_m$ is the melting temperature of the material) \cite{Chiou-2005, Lee-2005, Lee-2006, Chen-2007}.
Moreover, this model, in its original form, does not take into account the coupling between deformation and temperature effects \cite{Samantaray-2009}.
In order to reduce these shortcomings (and to apply this model to a wider range of temperatures) of the ZA model, modified forms of Zerilli-Armstrong (MZA) have been proposed by several authors in recent years \cite{NematNasser-2004, Lennon-2004, Muralli-2017, Cheng-2021, Muralli-2021}. The Arrhenius-type hyperbolic sine constitutive model is one of the most widely used phenomenological models both for the prediction of the hot deformation behavior of materials and for the analysis of the micro-structure of the material under study because this model allows the estimation of the activation energy for deformation \cite{Jonas-1969, Mostafaei-2012, Zhang-2012}.
Its original formulation has been revised several times for a more accurate description of the high temperature deformation behavior. In order to increase the accuracy of the model, the effects of strain \cite{Slooff-2007, Li-2012, Xu-2013}, strain rate \cite{Lin-2008-C, Mandal-2009} have been considered by several authors. For example, Sloof \eal \cite{Slooff-2007} introduced a strain-dependent parameter into the initial formulation of the Arrhenius model so that his model could predict the high-temperature flow of a magnesium alloy. Lin \eal \cite{Lin-2008-C}, on the other hand, modified the model by introducing a strain rate compensation factor. Another form of model developed by Hansel and Spittel \cite{Hensel-1978}, commonly named as the HS model, which is a thermo-visco-plastic model whose formulation is based on a product of functions dependent on $\varepsilon$, $\mdot\varepsilon$ and $T$. Unlike the JC model, this formulation is based on parameters that are determined simultaneously. This model is simple to use and has 8 parameters that are easy to determine. The limitation of this model is its loss of accuracy in some cases when the 8 parameters are defined.
This is why many works often limit its identification to only 5, 6 or 7 parameters}.

The fixed and limited number of parameters imposed by each analytical model as presented above has both advantages and disadvantages.
These models have few parameters and are therefore easy to identify, but when these models have to be applied to a material different from the one for which they were initially developed they lose their performance as in the case of the HS model.
To try to solve this problem, Tize Mha \eal \cite{TizeMha-2022} recently proposed a formulation where the number of parameters is not known in advance. The idea is to make the formulation of an analytical model flexible so that it can be adapted to many materials. This formulation is based on the MZA model and polynomial functions of undefined order are used during the identification but the problem with this model is the high order of the polynomial function that can affect its accuracy. Therefore a good choice of order of polynomial functions is important during its identification. Another identification technique is that based on artificial intelligence.

Artificial Neural Network (ANN) is an approach to predict the flow stress of materials without requiring a mathematical formulation of the flow law. It is therefore not necessary to postulate a mathematical expression to identify the parameters of the model. Since the behavior of materials is highly nonlinear at high temperatures, and depends on many factors that are also nonlinear, the evaluation of the flow stress by an analytical model whose parameters are identified by a classical regression method is limited. Faced with these limitations, ANN models are of major interest because they are particularly suited to deal with complex and nonlinear relationships. Consequently, ANNs have been successfully applied to predict the flow of materials under hot working conditions \cite{Lin-2008-ANN, Lu-2011-ANN, Ashtiani-2016-CSP, Stoffel-2019-NNB, Pantale-2021}. They provide a fundamentally different approach to statistical or numerical techniques for modeling materials.

\textcolor{red}{From all the work previously presented on both analytical and ANN models, a remark on the performance of these models emerges. Indeed, a model is considered appropriate to describe the behavior of a material when its predictions and experience correlate well. First an experiment is carried out, then the data from this experiment are used to determine the parameters of the model, and based on these parameters the predictions of this model and the initial data are displayed to see the correlation between predictions and experience. It is questionable whether a model that correctly describes the behavior of a material can be used to generate data for conditions different from those on which it was identified ? And in the event that we use it, how can we reassure ourselves of the relevance of what it predicts ? To answer these questions the interpolation and extrapolation techniques \cite{wahab2017interpolation,luo2019interpolation,stinis2019enforcing} widely used in several fields are extended to the constitutive laws in this work where the variable is the strain rate and this can be applied as well as to strain to the temperature. It is a question of choosing a strain rate different from those used during the identification of the model both inside and outside the range of known strain rates. An experiment is therefore carried out for this strain rate in order to validate the prediction of the models.}

In this work, the high-temperature thermomechanical behavior of medium carbon steel alloy subjected to compression tests in a Gleeble thermomechanical simulator was studied. From the experimental data obtained from the compression tests, five analytical models were identified in order to select the one that best describes the behavior of this alloy. A flow model based on an ANN was also developed and identified, and a comparative study between the 5 analytical models and the ANN model is proposed.
As presented previously a fundamental concern addressed in this work is the ability of a model to interpolate and extrapolate the results of the model with respect to the range of conditions covered by the experimental tests.

@@

The material used in this study is a \textcolor{red}{medium carbon steel} whose chemical composition is given in Table \ref{tab:Composition}.
\begin{table}[h!]
\centering
\caption{Chemical composition of medium carbon steel, Fe = balance.}
\begin{tabular}{lccccccc}
	\hline
	Element &  C   &  Mn  &  Mo  &  Si  &  Ni  &  Cr  &  Cu  \\ \hline
	Wt~\%   & 0.30 & 0.89 & 0.52 & 0.34 & 0.68 & 1.86 & 0.17 \\ \hline
\end{tabular}
\label{tab:Composition}
\end{table}
Cylindrical samples were machined with an initial diameter of $d=10$~mm and a height of $h_0=15$~mm.
\textcolor{red}{Hot compression} tests were performed on a Gleeble-3800 thermomechanical simulator (see Figure \ref{fig:Gleeble3800}), for the 5 temperature levels of $1050\celsius$, $1100\celsius$, $1150\celsius$, $1200\celsius$ and $1250\celsius$ with the 6 strain rates of $0.001~\ps$, $0.01~\ps$, $0.1~\ps$, $1~\ps$, $2.0~\ps$ and $5~\ps$.
\begin{figure}[!ht]
\centering
\includegraphics[width=0.7\columnwidth]{Figures/Gleeble-3800}
\caption{The Gleeble-3800 thermomechanical simulator system used for this study}
\label{fig:Gleeble3800}
\end{figure}
Thin tantalum sheets were used as a lubricating material at the contact surface of the anvils and samples to minimize friction during testing. As shown in Figure \ref{fig:GleebleProcess}, the samples were heated to a temperature of $1260\celsius$ with a heating rate of $2$\celsius/s and held at this temperature for $5$~min to eliminate thermal gradients. They were then cooled down with a rate of $1$\celsius/s to the test temperature and then held at constant temperature for $1$~min before deformation. During the compression phase, the temperature of the specimen is kept constant by the thermal control system of the machine. After compression, the specimen is quickly quenched to freeze its microstructure for later analysis. Figure \ref{fig:GleebleProcess} also shows the aspects of the specimens before and after the compression test: $h_0$ and $r_0$ are the height and radius before compression and $h$, $r_m$ and $r_t$ are the height, large radius and small radius of the specimen after compression, respectively.
\begin{figure}[!ht]
\centering
\includegraphics[width=0.8\columnwidth]{Figures/GleebleProcess}
\caption{Schematic diagram of the experimental process.}
\label{fig:GleebleProcess}
\end{figure}

>>>>

\title{Interpolation and extrapolation based performance measurement of Analytical and Artificial Neural Network models in predicting hot deformation behavior of a medium carbon steel}

@@

\begin{abstract}
\textcolor{red}{In the present work, a critical analysis of the most commonly used analytical models as well as recently introduced Artificial Neural Network based models is conducted to evaluate their prediction accuracy within and outside the experimental interval used to generate them}. To this end, the high-temperature deformation behavior of a medium carbon steel was studied over a wide range of strains ($0.0-0.7$), strain rates ($0.001~\ps-5~\ps$) and temperatures ($1050\celsius-1250\celsius$) using hot compression tests on a Geleeble $3800$ thermomechanical simulator. The experimental flow curves were then modeled using Johnson-Cook, Modified Zerilli-Armstrong, Hansel-Spittel, Arrhenius, and the recently proposed PTM model, as well as an Artificial Neural Network model with $3$ inputs, $2$ hidden layers of $15$ and $7$ neurons, respectively, and 1 output. The Mean Absolute Relative Error (MARE) and the Root Mean Square Error (RMSE) values were used to quantify the predictive accuracy of the analyzed models. The results indicated that the Johnson-Cook and Modified Zerilli-Armstrong models presented significant error, MARRE of  $21.20\%$, while the Hansel-Spittel, PTM and Arrhenius models could predict the behavior of this alloy with about $3.56\%$ error. The Artificial Neural Network model, on the other hand, showed excellent agreement, less than $0.62\%$ error, between the predicted and experimental flow curves. To further validate the performance of the above models, their ability to interpolate and extrapolate the experimental data was also tested. The Hansel-Spittel, PTM and Arrhenius analytical models showed a good interpolation and extrapolation capability with a notable superiority for the Arrhenius model compared to the other two. The Artificial Neural Networks model was however, still the most performant compared to all the models. A correlation factor is proposed to extend the applicability of some of the investigated models to deformation conditions not included in the original experimental plan.
\end{abstract}

@@

%----------------------------------------------------------------------------------
\section{Introduction\label{sec:Introduction}}
%----------------------------------------------------------------------------------
The manufacturing of large size forged blocks made of high-strength steels (as medium carbon alloy steel) is a process whose first step is ingot casting, followed by forging, quenching, and tempering operations. These forged blocks are used in automative industry  as die materials for the production of plastic injection components \cite{chadha2017deformation,chadha2018influence,murugesan2019two, murugesan2019hybrid}. For more industrial applications, requirements such as chemical homogeneity, uniform hardness and grain size inside these large size forged blocks are importants. The ingot breakdown which is the first step of the deformation process, is carried out generally to get the above requirements and strengthening or weakening of alloy may result. Work hardening (WH), phase transformation, and precipitation are those that often have strengthening effects while dynamic recovery (DRV), metadynamic recristallization (MDRX), dynamic recristallization (DRX), static recovery and static recristallization occur in the material softening. Strengthening or weakening effects strongly depends on three macroscopic quantities \ie the strain $\varepsilon$, the strain rate $\mdot\varepsilon$ and the temperature $T$ and impact the mechanical properties of the final product \cite{tavakoli2019ferrite, ebrahimi2017flow,Ashtiani-2012,He-2013, Changizian-2012}. As conducting an experiment is very expensive and time consuming, numerical simulation such as the Finite Element Analysis (FEA), where a good result depends on the behavior laws of the materials, is widely used nowadays to reduce experimental errors \cite{Mandal-2009, Qin-2010, Ji-2018,Lin-2008-P, Wu-2012}.

To predict the flow stress behavior during the breakdown process of ingots, many constitutive models have been proposed to simulate the thermomechanical process among which, phenomenological, semi-empirical, physical models or models based on Artificial Neural Networks (ANN) \cite{Rusinek-2010, Shin-2010, Lin-2011, Pantale-2021}. Among the large number of models available in the literature, the Johnson-Cook (JC) \cite{Johnson-1983} and Zerilli-Armstrong (ZA) \cite{Zerilli-1987} models are the best known and most widely available in FEA codes. The JC model is the most widely used because it is simple to identify and use and has few parameters to determine \cite{NematNasser-2003, Khan-2004}. However, this model, in its original state, has some shortcomings due to non-coupling effect of strain, strain rate and temperature as analyzed by Jia \eal \cite{Jia-2021}. To circumvent these shortcomings, several modified forms of the JC model have been developed \cite{Rule-1998, Vural-2003, Lin-2010, Lin-2012, Li-2013, Zhang-2015, Zhou-2019}.
Due to its more physics-based formulation, the ZA model is often preferred to the JC model because it allows the coupling of strain rate and temperature effects \cite{Johnson-1988, Voyiadjis-2005, Dey-2007}.
However, this model has other drawbacks related to the limit value of the test temperature. Indeed, this model applies only to low strain rates and for temperatures limited to about 60~\% of $T_m$ (where $T_m$ is the melting temperature of the material) \cite{Chiou-2005, Lee-2005, Lee-2006, Chen-2007}.
Moreover, the ZA model, in its original form, does not take into account the coupling between deformation and temperature effects \cite{Samantaray-2009}.
In order to reduce these shortcomings (and to apply this model to a wider range of temperatures) of the ZA model, modified forms of Zerilli-Armstrong (MZA) have been proposed \cite{NematNasser-2004, Lennon-2004, Muralli-2017, Cheng-2021, Muralli-2021}. The Arrhenius-type hyperbolic sine constitutive model is one of the most widely used phenomenological models both for the prediction of the hot deformation behavior of materials and for the analysis of the micro-structure of the material under study because this model allows the estimation of the activation energy for deformation \cite{Jonas-1969, Mostafaei-2012, Zhang-2012}.
Its original formulation has been revised several times for a more accurate description of the high temperature deformation behavior. In order to increase the accuracy of the model, the effects of strain \cite{Slooff-2007, Li-2012, Xu-2013}, strain rate \cite{Lin-2008-C, Mandal-2009} have been considered by several authors. For example, Sloof \eal \cite{Slooff-2007} introduced a strain-dependent parameter into the initial formulation of the Arrhenius model so that his model could predict the high-temperature flow of a magnesium alloy. Lin \eal \cite{Lin-2008-C}, on the other hand, modified the model by introducing a strain rate compensation factor. Another form of model is developed by Hansel and Spittel \cite{Hensel-1978} whose formulation is based on a product of functions dependent on $\varepsilon$, $\mdot\varepsilon$ and $T$. This model is simple to use and has 8 parameters that are easy to determine. The limitation of this model is its loss of accuracy in some cases when the 8 parameters are defined.
This is why many works often limit its identification to only 5, 6 or 7 parameters.

Most of the developed models are related to a specific material or alloy and where they are used for other materials, their efficiency is not good. To overcome this issue, Tize Mha \eal \cite{TizeMha-2022} recently proposed a constitutive model whose formulation can be adapt to any kind of material or alloys. This formulation is based on the MZA model and polynomial functions of undefined order are used during the identification but the problem with this model is the high order of the polynomial function that can affect its accuracy. Therefore a good choice of order of polynomial functions is important during its identification. Another model is that based on artificial intelligence method where parameters identification step of analytical models is replaced by training step.

Artificial Neural Network (ANN) is an approach to predict the flow stress behavior of materials without requiring a mathematical formulation of the flow law. It is therefore not necessary to postulate a mathematical expression to identify the parameters of the model. Since the behavior of materials is highly nonlinear at high temperatures, and depends on many factors that are also nonlinear, the evaluation of the flow stress behavior by an analytical model whose parameters are identified by a classical regression method is limited. Faced with these limitations, ANN models are of major interest because they are particularly suited to deal with complex and nonlinear relationships. Consequently, ANNs have been successfully applied to predict the flow stress behavio of materials under hot working conditions \cite{Lin-2008-ANN, Lu-2011-ANN, Ashtiani-2016-CSP, Stoffel-2019-NNB, Pantale-2021}.

Although progress has been made in improvement of constitutive models to better predict the flow stress behavior of materials, there is still a problem with the efficiency of the model. Indeed, a model is considered appropriate to the prediction of the material's behavior when its predictions and experience correlate well. It is questionable whether a model that correctly describes the behavior of a material can be used to generate data for conditions different from those on which it was identified and how to know the relevance of what it predicts. Interpolation and extrapolation techniques \cite{wahab2017interpolation,luo2019interpolation,stinis2019enforcing} widely used in several fields are extended to the constitutive models in this work where the variable is the strain rate and this can be applied as well as to strain to the temperature. It is a question of choosing a strain rate different from those used during the identification of the model both inside and outside the range of known strain rates. An experiment is therefore carried out for this strain rate in order to validate the prediction of the models. For this purpose, the high-temperature thermomechanical behavior of medium carbon steel alloy subjected to compression tests in a Gleeble thermomechanical simulator was studied. From the experimental data, five analytical models were identified in order to select the one that best describes the behavior of the studied material. A flow model based on an ANN was also developed and identified, and a comparative study between the 5 analytical models and the ANN model is proposed. As presented previously a fundamental concern addressed in this work is the ability of a model to interpolate and extrapolate the results of the model with respect to the range of conditions covered by the experimental tests.

@@

The material used in this study is a medium carbon steel whose chemical composition is given in Table \ref{tab:Composition}.
\begin{table}[h!]
\centering
\caption{Chemical composition of medium carbon steel, Fe = balance.}
\begin{tabular}{lccccccc}
	\hline
	Element &  C   &  Mn  &  Mo  &  Si  &  Ni  &  Cr  &  Cu  \\ \hline
	Wt~\%   & 0.30 & 0.89 & 0.52 & 0.34 & 0.68 & 1.86 & 0.17 \\ \hline
\end{tabular}
\label{tab:Composition}
\end{table}
Cylindrical samples were machined with an initial diameter of $d=10$~mm and a height of $h_0=15$~mm.
Hot compression tests were performed on a Gleeble-3800 thermomechanical simulator (see Figure \ref{fig:Gleeble3800}), for the 5 temperature levels of $1050\celsius$, $1100\celsius$, $1150\celsius$, $1200\celsius$ and $1250\celsius$ with the 6 strain rates of $0.001~\ps$, $0.01~\ps$, $0.1~\ps$, $1~\ps$, $2.0~\ps$ and $5~\ps$.
\begin{figure}[!ht]
\centering
\includegraphics[width=0.7\columnwidth]{Figures/Gleeble-3800}
\caption{The Gleeble-3800 thermomechanical simulator system used for this study}
\label{fig:Gleeble3800}
\end{figure}
Thin tantalum sheets were used as a lubricating material at the contact surface of the anvils and samples to minimize friction during testing. As shown in Figure \ref{fig:GleebleProcess}, the samples were heated to a temperature of $1260\celsius$ with a heating rate of $2$\celsius/s and held at this temperature for $5$~min to eliminate thermal gradients. They were then cooled down with a rate of $1$\celsius/s to the test temperature and then held at constant temperature for $1$~min before deformation. During the compression phase, the temperature of the specimen is kept constant by the thermal control system of the machine. After compression, the specimen is quickly quenched to freeze its microstructure for later analysis. Figure \ref{fig:GleebleProcess} also shows the aspects of the specimens before and after the compression test: $h_0$ and $r_0$ are the height and radius before compression and $h$, $r_m$ and $r_t$ are the height, large radius and small radius of the specimen after compression, respectively.
\begin{figure}[!ht]
\centering
\includegraphics[width=0.8\columnwidth]{Figures/GleebleProcess}
\caption{Schematic diagram of the experimental process.}
\label{fig:GleebleProcess}
\end{figure}
